---
title: "Wine Project - Part C"
author: "Kowsalya Muralidharan"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

1.  **Produce summary statistics of "residual.sugar" and use its median
    to divide the data into two groups A and B. We want to test if
    "density" in Group A and Group B has the same population mean.
    Please answer the following questions.**

```{r}
Red_Wine_Data<-read.csv(file = "C:/Users/kowsa/OneDrive/Desktop/UC/ACADS/Applied Stat Methods/Wine Project/winequality-red.csv",header = T, sep=";")
attach(Red_Wine_Data)

summary(Red_Wine_Data$residual.sugar)
```

```{r}

residual.sugar.group<-ifelse(residual.sugar<2.20, 0, 1)
data.frame(residual.sugar, residual.sugar.group, density)[1:10,]
```

a.  **State the null hypothesis;**

Null Hypothesis (H0): The population means of "density" in Group A and
Group B are equal.

H0: μA = μB

b.  **Use visualization tools to inspect the hypothesis. Do you think
    the hypothesis is right or not?**

```{r}
boxplot(density ~ residual.sugar.group)
```

From the boxplot, we can see that there are differences in the means of
"density" among the groups. This indicated that there could be evidence
against the null hypothesis. However, to conclude we need to perform
statistical testing.

c.  **What test are you going to use?**

We can use T-Test to test the hypothesis.

```{r}
t.test(density ~ residual.sugar.group)
```

d.  **What is the p-value?**

The p-value is \< 2.2e-16, which is very negligible compared to the
threshold of 0.05.

e.  **What is your conclusion?**

I suggest that we reject the null hypothesis, indicating that there is
evidence that the population means of "density" in the two groups are
different. However, there is a 5% chance that we may be wrong about
rejecting the hypothesis when it is true.

f.  **Does your conclusion imply that there is an association between
    "density" and "residual.sugar"?**

Since we rejected the null hypothesis, we can say there is an
association between density and residual sugar.

2.  **Produce summary statistics of "residual.sugar" and use its 1st,
    2nd, and 3rd quantiles to divide the data into four groups A, B, C,
    and D. We want to test if "density" in the four groups has the same
    population mean. Please answer the following questions.**

```{r}
summary(residual.sugar)

residual.sugar.group2<-NULL

for(i in 1:1599){
if(residual.sugar[i]<=1.9) residual.sugar.group2[i]<-1
else if(residual.sugar[i]<=2.20) residual.sugar.group2[i]<-2
else if(residual.sugar[i]<=2.60) residual.sugar.group2[i]<-3
else residual.sugar.group2[i]<-4
}
table(residual.sugar.group2)
```

a.  **State the null hypothesis;**

The null hypothesis (H0) is that the population means of "density" in
all four groups are equal:

H0: μA = μB = μC = μD

b.  **Use visualization tools to inspect the hypothesis. Do you think
    the hypothesis is right or not?**

```{r}
boxplot(density ~ residual.sugar.group2)
```

From the boxplot, we can see that there are differences in the means of
"density" among the groups. This indicated that there could be evidence
against the null hypothesis.

c.  **What test are you going to use?**

To test if the population means of "density" in the four groups are the
same, we can use ANOVA test.

```{r}
summary(aov(density ~ residual.sugar.group2))
```

d.  **What is the p-value?**

The p-value is \<2e-16, which is negligible compared to the threshold of
0.05.

e.  **What is your conclusion?**

A very small p-value means that there are significant differences in the
means of the groups with context to residual.sugar.

Given this very small p-value, we would typically reject the null
hypothesis (H0) that the population means of "density" are the same
across the groups.

f.  **Does your conclusion imply that there is an association between
    "density" and "residual.sugar"? Compare your result here with that
    in Question 1. Do you think increasing the number of groups help
    identify the association? Would you consider dividing the data into
    10 groups so as to help the discovery of the association? Why?**

By rejecting the null hypothesis, we can conclude that there are
differences in means among the groups, and that there is association
between "density" and "residual.sugar." This was similar to the case in
Question 1, and both the p-values are close.

Increasing the data into 10 groups will help with figuring out more
nuanced differences in means within different segments of the data, but
it will still point to the same conclusion derived above.

3.  **Create a 2 by 4 contingency table using the categories A, B, C, D
    of "residual.sugar" and the binary variable "excellent" you created
    in Part B. Note that you have two factors: the categorical levels of
    "residual.sugar" (A, B, C and D) and an indicator of excellent
    wines(yes or no).**

```{r}
Red_Wine_Data$excellent<- ifelse(Red_Wine_Data$quality>=7,1,0)

contingency_table <- table(residual.sugar.group2, Red_Wine_Data$excellent)

# Rename rows and columns for clarity
rownames(contingency_table) <- c("A", "B", "C", "D")
colnames(contingency_table) <- c("Not Excellent", "Excellent")

# Print the contingency table
contingency_table
```

a.  **Use the Chi-square test to test if these two factors are
    correlated or not;**

```{r}
Xsq <- chisq.test(contingency_table)
Xsq$observed # observed counts (same as M)
Xsq$expected # expected counts under the null
Xsq$residuals # Pearson residuals

Xsq
```

Here, the p-value is 0.1386 \> 0.05 (threshold), which means that there
is no strong evidence to overthrow the hypothesis, suggesting that there
is no significant association.

b.  **Use the permutation test to do the same and compare the result to
    that in (a);**

```{r}
chisq.test(contingency_table,simulate.p.value = T)
```

Here, we can see that the p-value is 0.1409, which is also greater than
the threshold 0.05. There is no strong statistical evidence to conclude
that there is a significant association between "residual.sugar"
categories and the "excellent" wines.

c.  **Can you conclude that "residual.sugar" is a significant factor
    contributing to the excellence of wine? Why?**

No, we cannot conclude that residual sugar is a significant factor as
both in the chi-squared and permutation test, we got a p-value \> 0.05,
so we do not have enough evidence to suggest that the two factors are
correlated or associated with each other. In other words, the null
hypothesis is true.
